Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-29 Manuscript/manuscript.Rmd
reordered figures

## TAD boundary data

@@

## Genomic annotation data

Genomic annotation data in the form of cell type specific chromatin states (ChromHMM), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 2). [??? The former two may be combined, TBD]

@@

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]), we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, and 9314, 10881, 6923, 3197, and 831 unique TAD boundaries for the K562 cell line across each of the five resolutions respectively (Supplementary Figure 1). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). On average, we found that only $18.0 \pm 4.6 \%$ of TAD boundary regions overlapped between both cell lines (Supplementary Figure 2). Given the cell type specificity of TAD boundaries, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. In summary, we developed a framework enabling resolution-specific prediction of TAD boundaries, with resolution-specific bins defining the response variable.

@@

A total of 78 cell type specific functional genomic annotations for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**overlap count**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figure 4). These feature engineering steps were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

@@

Given that the number of bins annotated as TAD boundaries (TAD boundary regions) is much smaller than the number of non-TAD boundary bins, we investigated the class imbalance (CI) featured at each resolution. The CI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions and ranging from 0 to 0.50. The CI was found to be invariant to resolution, remaining consistently low across chromosomes, with an average of $0.196 \pm 0.025$ and $0.166 \pm 0.026$ for the GM12878 and K562 cell lines, respectively (Supplementary Figure 3). To account for such imbalance, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

@@

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using one of the predictor type *P = {OC, OP, Distance}*, and the re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 10-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 = 2520$) models. Additional performance metrics are reported in the Supplementary materials.

@@

We investigated the effect of class imbalance on model performance using three re-sampling techniques. Expectedly, when using data as-is, that is, with class imbalance present, the models performed poorly. Using any of the resampling techniques improved the performance, with SMOTE-resampled data being the most optimal (Figure 3, Supplementary Figure 3). These results suggest that a combination of both random over- and under-sampling (SMOTE) can improve TAD boundary prediction.

We found that, regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the overlap count and overlap percent predictor types (Figure 3), and these results were consistent across resolutions and cell types (Supplementary Figure 5). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.

Interestingly, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be due to the sparsity in the interaction frequency matrices at finer resolutions. [??? Quantify sparsity, need to substantiate this statement; !!!divide nonzero elements by total?] This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Additionally, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. In summary, our results imply that using random under- and over-sampling, distance-type predictors, and the data at 10kb resolution results in an optimally performing cell line-specific random forest-based TAD boundary prediction model.

@@

We then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the three groups of genomic annotations (histone modifications (HM), chromatin states (ChromHMM), and transcription factor binding sites (TFBS)). For optimal random forest classifiers, we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites (Supplementary Figure 6). Furthermore, models build solely on TFBSs outperformed other annotation-specific models at all resolutions (Figure 4A; Supplementary Figure 7A). Notably, the TFBS-specific models outperformed models built on the combined sets of annotations, and this effect was most pronounced at lower resolutions. This indicated that the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction. In summary, these results suggest that TFBS are the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated (Supplementary Figure 8), we sought to avoid overfitting by performing recursive feature elimination. We found that our models were able to obtain near-optimal performance using approximately eight TFBSs (Figure 4B; Supplementary Figure 7B). However, given that we train our models on chromosome-specific data, the most significant annotations vary for each chromosome. To determine the most common TFBSs, we clustered the predictive importance values across chromosomes (Figure 4C; Supplementary Figure 7C). We found that, similar to other studies [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions.

@@

After obtaining our optimally built random forest classifier, we investigated whether our predictive models can alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. We were able to identify regions where the model predicted the base pair to be a TAD boundary with probability equal to 1 (Figure 5A). [??? Where is this figure? It seemed important; !!!added] By clustering the regions (see Methods), we were able to predict a comparable number of TAD boundary regions as called by the ARROWHEAD TAD caller. However, only 24% and 29% of predicted and called (and 10 kb flanked) TAD boundaries overlapped for the GM12878 and K562 cell line, respectively, suggesting differing locations in TAD boundaries throughout the linear genome (Supplementary Figure 9). [??? Need further drill down, as we discussed; !!!added distance to annotations figure in next paragraph]

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!; !!!fixed] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5B; Supplementary Figure 10).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs. Furthermore, when compared to called TAD boundaries, the boundaries that we predicted with our model were found to be statistically significantly closer to the top ranked TFBS, suggesting stronger enrichment profiles near predicted TAD boundaries (Supplementary Figure 11).

@@

**Supplementary Table 1. Hi-C data.** A complete list of the Hi-C data libraries used to obtain the locations of topologically associating domains.

**Supplementary Table 2. Genomic annotations.** A complete list of the cell line specific genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic annotations, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Figure 1. Diagram of the number of called TADs and TAD boundaries.** TADs were called using the ARROWHEAD pipeline [@durand2016juicebox] at 5 resolutions: 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb, for both the GM12878 and K562 cell line, across each autosomal chromosome (excluding chromosome 9). A preprocessing step was included to filter the TADs that exceeded 2 mb in width. The coordinates of unique TAD boundaries were then extracted and enumerated at each resolution.

**Supplementary Figure 2. Venn diagrams showing that TAD boundary regions exhibit cell-line specificity across resolution.**

**Supplementary Figure 3. Boxplots illustrating that the cell line specific class imbalance remained severe irrespective of resolution.** Class imbalance (CI) was defined as the ratio of TAD boundary regions to the sum of both TAD and non-TAD boundary regions. The boxplots provide summaries of CI across chromosomes across resolution, for the GM12878 (A) and K562 (B) cell lines.

**Supplementary Figure 4. Performing $log_{2}$ transformations normalizes skewed distance densities across resolution.** When measured in basepairs, distances were highly right skewed (red lines) across all resolutions for both the GM12878 (A) and K562 (B) cell lines. After $log_{2}$ transforming the distances (blue lines), densities became much more gaussian distributed for all resolutions.

**Supplementary Figure 5. Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal MCCs across resolution on K562.** Barplots comparing the performances of random forest classifiers built using different combinations of data level characteristics for K562. Within each re-sampling technique (A-D), performances are compared for each type of feature space: OC (red), OP (green), and Distance (blue), across resolutions. Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Supplementary Figure 6. The proportional importance of optimally built cell line specific models was outweighed by transcription factor binding sites across resolutions.** Stacked barplots illustrate the proportion of total predictive importance in cell line specific random forest classifiers for (A) GM12878 and (B) K562 respectively. Model inputs included the optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors). Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS).

**Supplementary Figure 7. Assessing top ranked transcription factors in TAD boundary prediction for K562.** (A) Barplots comparing performances of models using K562-specific groups of functional genomic annotation as the feature space including histone modifications (HM; red), chromatin states (ChromHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all 3 groups (Combined; purple). Error bars represent ± 1 standard deviation from the mean MCC across each chromosomal section of the linear genome. (B) Recursive feature elimination analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. Error bars represent 1 standard deviation away from the mean cross-validated MCC. (C) Clustered heatmap of predictive importance for the union of the top 8 most predictive TFBS across chromosomes.

**Supplementary Figure 8. Pairwise Jaccard Indices illustrate correlations between cell line specific transcription factor binding sites for GM12878 (A) and K562 (B).**

**Supplementary Figure 9. Venn diagrams illustrating the amount of overlap between predicted and called TAD boundary regions.** For each of the GM12878 (A) and K562 (B), both predicted and called TAD boundaries were flanked other either side by 10 kb and the amount of overlap was enumerated.

**Supplementary Figure 10. Predicted TAD boundaries were centered closer to peak signal regions defined by known architectural transcription factors when compared to called TAD boundaries for K562.** Enrichment
heatmaps illustrate the average signal strength of K562-specific ChIP-seq peak regions defined by (A) SMC3, (B) ZNF143, (C) CTCF, and (D) RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.

**Supplementary Figure 11. Known architectural transcription factors were statistically significantly closer to predicted TAD boundaries then called TAD boundaries.** Barplots illustrate the log2 distance in basepairs from classes of TAD boundaries to transcription factors including CTCF, RAD21, SMC3, and ZNF143 for the GM12878 (A) and K562 (B) cell lines. TAD boundaries were classified into the 3 classes. Predicted but not called (PBNC) were the set of 10 kb flanked predicted TAD boundaries the did not overlap with 10 kb flanked called TAD boundaries. Called and predicted (CAP) were the set of 10 kb flanked called and predicted TAD boundaries that overlapped with each other. Called but not predicted (CBNP) were the set of 10 kb flanked called TAD boundaries that did not overlap with 10 kb flanked predicted TAD boundaries.

>>>>

## Genomic annotation and TAD boundary data

@@

Genomic annotation data in the form of cell type specific chromatin states (ChromHMM), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 1).

@@

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]), we obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, across each of the five resolutions respectively (Table 1). Similar numbers of TAD boundaries were found for the K562 cell line (not reported). To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). On average, we found that only $18.0 \pm 4.6 \%$ of TAD boundary regions overlapped between both cell lines. Given the cell type specificity of TAD boundaries, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. In summary, we developed a framework enabling resolution-specific prediction of TAD boundaries, with resolution-specific bins defining the response variable.

| Resolution | Total number of called TADs | Total number of unique TAD boundaries | Average Class Imbalance (SD) |
|:----------:|:---------------------------:|:-------------------------------------:|:----------------------------:|
|5 kb        |8043	                       | 16133                                 | 0.06 (0.02)                  |
|10 kb	     |7650	                       | 14824                                 | 0.10 (0.01)                  |
|25 kb	     |4208	                       | 8081                                  | 0.14 (0.01)                  |   
|50 kb	     |1772	                       | 3560	                                 | 0.12 (0.01)                  |
|100 kb	     |444	                         | 919	                                 | 0.06 (0.01)                  |    
Table: Domain data and class imbalance summaries across resolutions for GM12878.

A total of 78 cell type specific functional genomic annotations for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Table 1). The annotations could be broken down into three sets including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**overlap count**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figure 1). These feature engineering steps were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

@@

@@

Given that the number of bins annotated as TAD boundaries (TAD boundary regions) is much smaller than the number of non-TAD boundary bins, we investigated the class imbalance (CI) featured at each resolution. The CI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions and ranging from 0 to 0.50. The CI was found to be invariant to resolution, remaining consistently low across chromosomes, with an averages of 0.06, 0.10, 0.14, 0.12, and 0.06 across each resolutions for GM12878 (Table 1). Similar class imbalances were found for the K562 cell line (not reported). To account for such imbalance, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

@@

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using one of the predictor type *P = {OC, OP, Distance}*, and the re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 10-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 = 2520$) models. <!--Additional performance metrics are reported in the Supplementary materials.-->

@@

We investigated the effect of class imbalance on model performance using three re-sampling techniques. Expectedly, when using data as-is, that is, with class imbalance present, the models performed poorly. Using any of the resampling techniques improved the performance, with SMOTE-resampled data being the most optimal (Figure 3, Supplementary Figure 2). These results suggest that a combination of both random over- and under-sampling (SMOTE) can improve TAD boundary prediction.

We found that, regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the overlap count and overlap percent predictor types (Figure 3), and these results were consistent across resolutions and cell types (Supplementary Figure 2). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations.

Interestingly, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be due to the sparsity in the interaction frequency matrices at finer resolutions. [??? Quantify sparsity] This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Additionally, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. In summary, our results imply that using random under- and over-sampling, distance-type predictors, and the data at 10kb resolution results in an optimally performing cell line-specific random forest-based TAD boundary prediction model.

@@

We then aimed to compare the performances of annotation-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the three groups of genomic annotations (histone modifications (HM), chromatin states (ChromHMM), and transcription factor binding sites (TFBS)). For optimal random forest classifiers, we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites (Supplementary Figure 3). Furthermore, models build solely on TFBSs outperformed other annotation-specific models at all resolutions (Figure 4A; Supplementary Figure 4A). Notably, the TFBS-specific models outperformed models built on the combined sets of annotations, and this effect was most pronounced at lower resolutions. This indicated that the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction. In summary, these results suggest that TFBS are the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated, we sought to avoid overfitting by performing recursive feature elimination. We found that our models were able to obtain near-optimal performance using approximately eight TFBSs (Figure 4B; Supplementary Figure 4B). However, given that we train our models on chromosome-specific data, the most significant annotations vary for each chromosome. To determine the most common TFBSs, we clustered the predictive importance values across chromosomes (Figure 4C; Supplementary Figure 4C). We found that, similar to other studies [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions.

@@

After obtaining our optimally built random forest classifier, we investigated whether our predictive models can alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. We were able to identify regions where the model predicted the base pair to be a TAD boundary with probability equal to 1 (Figure 5A). [??? Where is this figure? It seemed important; !!!added] By clustering the regions (see Methods), we were able to predict a comparable number of TAD boundary regions as called by the ARROWHEAD TAD caller. However, only 24% and 29% of predicted and called (and 10 kb flanked) TAD boundaries overlapped for the GM12878 and K562 cell line, respectively, suggesting differing locations in TAD boundaries throughout the linear genome (Supplementary Figure 5). [??? Need further drill down, as we discussed; !!!added distance to annotations figure in next paragraph]

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!; !!!fixed] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5B; Supplementary Figure 6).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs. Furthermore, when compared to called TAD boundaries, the boundaries that we predicted with our model were found to be statistically significantly closer to the top ranked TFBS, suggesting stronger enrichment profiles near predicted TAD boundaries (Supplementary Figure 7).

@@

**Supplementary Table 1. Hi-C and genomic annotation data.** (A) A complete list of the Hi-C data libraries used to obtain the locations of topologically associating domains. (B) A complete list of the cell line specific genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic annotations, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Figure 1. Performing $log_{2}$ transformations normalizes skewed distance densities across resolution.** When measured in basepairs, distances were highly right skewed (red lines) across all resolutions for both the GM12878 (A) and K562 (B) cell lines. After $log_{2}$ transforming the distances (blue lines), densities became much more gaussian distributed for all resolutions.

**Supplementary Figure 2. Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal MCCs across resolution on K562.** Barplots comparing the performances of random forest classifiers built using different combinations of data level characteristics for K562. Within each re-sampling technique (A-D), performances are compared for each type of feature space: OC (red), OP (green), and Distance (blue), across resolutions. Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Supplementary Figure 3. The proportional importance of optimally built cell line specific models was outweighed by transcription factor binding sites across resolutions.** Stacked barplots illustrate the proportion of total predictive importance in cell line specific random forest classifiers for (A) GM12878 and (B) K562 respectively. Model inputs included the optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors). Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS).

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD boundary prediction for K562.** (A) Barplots comparing performances of models using K562-specific groups of functional genomic annotation as the feature space including histone modifications (HM; red), chromatin states (ChromHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all 3 groups (Combined; purple). Error bars represent ± 1 standard deviation from the mean MCC across each chromosomal section of the linear genome. (B) Recursive feature elimination analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. Error bars represent 1 standard deviation away from the mean cross-validated MCC. (C) Clustered heatmap of predictive importance for the union of the top 8 most predictive TFBS across chromosomes.

**Supplementary Figure 5. Venn diagrams illustrating the amount of overlap between predicted and called TAD boundary regions.** For each of the GM12878 (A) and K562 (B), both predicted and called TAD boundaries were flanked other either side by 10 kb and the amount of overlap was enumerated.

**Supplementary Figure 6. Predicted TAD boundaries were centered closer to peak signal regions defined by known architectural transcription factors when compared to called TAD boundaries for K562.** Enrichment
heatmaps illustrate the average signal strength of K562-specific ChIP-seq peak regions defined by (A) SMC3, (B) ZNF143, (C) CTCF, and (D) RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.

**Supplementary Figure 7. Known architectural transcription factors were statistically significantly closer to predicted TAD boundaries then called TAD boundaries.** Barplots illustrate the log2 distance in basepairs from classes of TAD boundaries to transcription factors including CTCF, RAD21, SMC3, and ZNF143 for the GM12878 (A) and K562 (B) cell lines. TAD boundaries were classified into the 3 classes. Predicted but not called (PBNC) were the set of 10 kb flanked predicted TAD boundaries the did not overlap with 10 kb flanked called TAD boundaries. Called and predicted (CAP) were the set of 10 kb flanked called and predicted TAD boundaries that overlapped with each other. Called but not predicted (CBNP) were the set of 10 kb flanked called TAD boundaries that did not overlap with 10 kb flanked predicted TAD boundaries.
